--- 
title: "Introduction to `spector` (SEquence COverage Roughness)"
author: "Anas A Rana"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    citation_package: natbib
vignette: >
  %\VignetteIndexEntry{Introduction to spector}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, echo = FALSE, message = FALSE}

# Lodaing packages
library(dplyr)
library(tidyr)
library(stringr)
library(spector)
library(knitr)
library(ggplot2)

knitr::opts_chunk$set(collapse = T, comment = "#>")
options(tibble.print_min = 4L, tibble.print_max = 4L)

theme_anas <- theme_set(theme_bw())
theme_anas <- theme_update(
  axis.text.x = element_text(size = rel(0.9)),
  axis.text.y = element_text(size = rel(0.9)),
  axis.title.x = element_text(size = rel(1)),
  axis.title.y = element_text(angle = 90, size = rel(1)),
  axis.ticks = element_line(colour = "grey90"),
  legend.key = element_rect(colour = "grey99", size = 0.1),
  legend.background = element_blank(),
  plot.background = element_blank(),
  panel.background = element_blank(),
  panel.border = element_rect(fill = NA, colour = "grey50"),
  panel.grid.major = element_line(colour = "grey90", size = 0.2),
  panel.grid.minor = element_line(colour = "grey98", size = 0.5),
  strip.text = element_text(colour="#DBDBDB"),
  strip.background = element_rect(colour="black", fill='black', size = 0.2)
  )
```

# Background

Recent advances in high-throughput sequencing of cancer genomes has elucidated the landscape of genetic alterations in many tumor types, thus expanding molecular classification of cancer and facilitating routine whole genome sequencing in clinical oncology.

The quality of the obtained tumor genome is a function of many variables, such as tissue quality and quantity, tumor content within a sample, depth of sequencing, and effectiveness of the data processing pipeline. Tissue quality relates to the age and size of the tissue and a fixation protocol. The fresh or frozen (FF) tumor material generally yields high-quality tumor genomic DNA. However, in most cases archival specimens stored as formalin-fixed, paraffin-embedded (FFPE) tissue blocks. Formalin can fragment the genomic DNA and introduce chemical modifications during the fixation and storage process that, eventually, lead to false-positive mutation and copy-number alteration calls. As FFPE tissue currently comprise the most accessible clinical option, evaluation of which samples yield successful whole genome sequence is critically important for molecular diagnostics and production-level sequencing.

## Aim

The aim of this package is to obtain a metric for evaluation as a proxy for the quality of the sequenced genomic data. The metric should allow for comparison of whole genome sequence datasets (e.g. samples preserved with FFPE and FF protocols) as well as regions of particular concern.

## Method

This package uses wavelets to deconstruct the signal into frequency space and identify noisy regions in the genome. The algorithm simple segments the genome into regions of equal size and computes a metric for each region. This metric acts as a proxy for the quality of the signal in this region. Briefly it is a measure on how much of the signal is made up of noise. 

The default regions used in `spector` are based on the genome in a bottle (giab) project ([more details](http://jimb.stanford.edu/giab/)). The regions are further subset to identify reliable regions using [ReliableGenome](http://github.com/popitsch/wtchg-rg).

# First run

To run `spector` the first time all we need it is access to a bam file. There are two very basic bam files provided with the package. The files are `sample1.bam` and `sample2.bam`. To obtain the path of these files we can use the function `spector_sample()`. 

The best way to obtain QC results from spector is to use the function `spector_qc()`. With default settings the default region size $= 2^{13}$. We notice that the package provides a progress update while running, which is especially useful for larger files.

```{r}

bam_f <- spector_sample("sample1.bam")
results_qc <- spector_qc(f_bam = bam_f)

```

## Region size

We can specify region sizes when running `spector_qc()` using the `region_size` variable. This variable should be a power of $2$ if it isn't it is coerced to the largest power of $2$ which is smaller than `< region_size` inside the package.

```{r, message = FALSE}

results_qc <- spector_qc(f_bam = bam_f, region_size = 2^14)

```

## All files in folder

It is also possible to run spector on all bam files contained inside a folder. All we need to do is to set the variable `f_bam = ` as the path to the folder containing the bam files. We can also specify a custom bed file, we select one supplied with the package corresponding to the bam files in the package. 

The package will first compute chromosomal overlaps between the bed file and the bam file and only compute metric values for regions in chromosomes contained in both the bam and the bed file.

```{r}

bam_folder <- spector_sample("")
basic_path <- spector_sample("basic.bed")

print(bam_folder)

results_qc <- spector_qc(f_bam = bam_folder, f_bed = basic_path)

```

## Parameter file

The final way to pass bam files to spector is to use a parameter file. The file needs to have a very specific structure. 

```{r}
par_file <- spector_sample("sample_id_2.txt")

cat(readChar(par_file, 1e5))

```

The required format is a tab-delimited file with three columns. The first contains the path to the bam file relative to the current working directory, it is recommended to use the absolute path to avoid ambiguity. The second column contains a unique sample id. The third column contains any further description of the sample type. Any further columns will be ignored. There should be no header in the parameter file, if there is it should start with a `#`. To read this parameter file spector has the function `read_par_file()`. 

```{r}
read_par_file(par_file) %>% 
  kable()

```

# Results

The results of a `spector_qc` run are a [`tbl_df`](https://cran.rstudio.com/web/packages/dplyr/vignettes/introduction.html) object, which is part of the `dplyr` package. It contains an `id` variablec which encodes the genomic region in standard format the metric values and any further sample specific parameters passed to the method. 

```{r}
glimpse(results_qc)

```


## Compare results

A simple way to visualise the results is to plot comparative boxplots, as shown in this example. 

```{r}
results_qc %>% 
  ggplot(aes(x = id_bam, y = metric)) +
  geom_boxplot()

```

This is not always satisfactory depending on the application. For example in this example the samples contain different chromosomes. We could do a chromosome specific comparison to get a more fine-grained picture. 

```{r, fig.width = 10, fig.height = 5}
results_qc %>% 
  ggplot(aes(x = id_bam, y = metric, fill = chrom)) +
  geom_boxplot() +
  facet_grid(.~chrom)

```

In this case this shows us that the samples are different only in chromosome 1 and chromosome 2. This is because the sample bam files originate from the same source file with different sub-setting. 
